---
title: "My tips for working with ggplot2 in a RMarkdown document"
author: "Benjamin Louis"
date: '2020-06-21'
slug: en
categories:
  - configuration
  - tips
tags:
  - R
  - rmarkdown
  - ggplot2
banner: 'img/banners/rmarkdown-ggplot2.png'
type: 'blog'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.align = "center",
  out.width = "70%",
  fig.width = 7)
```


<!--Writing a report with [R Markdown](https://rmarkdown.rstudio.com/) combined with [`ggplot2`](https://ggplot2.tidyverse.org/) visualisation is a real advantage for reproducible data analysis. Features of both packages are highly flexible and you CAN always get what you want! But if you are just starting out, getting what you want can be cumbersome In this post, I share with you some tips found over time.-->

<!--more-->

# Context

Writing [R Markdown](https://rmarkdown.rstudio.com/) document makes possible to insert [R](https://www.r-project.org/) code and its results in a report with a choosen output format (HTML, PDF, Word). It is a real asset for analysis reproducibility as well as communication of methods and results. Doing daily data analysis, I usually deliver outputs in report and R Markdown naturally became an essential tool of my workflow.

Data analysis without data visualisation is like playing darts in the dark, there is a good chance you'll miss the ~~bullseye~~ point. You'll find quite a few R packages to build graphics but I have a preference for [`ggplot2`](https://ggplot2.tidyverse.org/) (I'm not alone!). Therefore, `ggplot2` graphics are often included in my R Markdown documents.

Features of both packages are highly flexible and you **CAN** always get what you want ! But if you are just starting out, getting what you want can be cumbersome. In this post, I share with you some tips found over time.

__Requirements__


1. I assume you have already made a graphic with `ggplot2` or at least seen some `ggplot2` code. If not, you can have a look at [this book](https://ggplot2-book.org/) freely available online. I will use a data visualisation of [Palmer penguins data](https://github.com/allisonhorst/penguins) recently included in a R package by [Allison Horst](https://twitter.com/allison_horst?s=20) (go see her [illustrations](https://github.com/allisonhorst/stats-illustrations) too !). I start with this graphic :

```{r echo = FALSE, message = FALSE, results = 'hide', cache = TRUE}
library(tidyverse)
library(janitor)
# Adelie penguin data from: https://doi.org/10.6073/pasta/abc50eed9138b75f54eaada0841b9b86
uri_adelie <- "https://portal.edirepository.org/nis/dataviewer?packageid=knb-lter-pal.219.3&entityid=002f3893385f710df69eeebe893144ff"
# Gentoo penguin data from: https://doi.org/10.6073/pasta/2b1cff60f81640f182433d23e68541ce
uri_gentoo <- "https://portal.edirepository.org/nis/dataviewer?packageid=knb-lter-pal.220.3&entityid=e03b43c924f226486f2f0ab6709d2381"
# Chinstrap penguin data from: https://doi.org/10.6073/pasta/409c808f8fc9899d02401bdb04580af7
uri_chinstrap <- "https://portal.edirepository.org/nis/dataviewer?packageid=knb-lter-pal.221.2&entityid=fe853aa8f7a59aa84cdd3197619ef462"
# Combining the URIs
uris <- c(uri_adelie, uri_gentoo, uri_chinstrap)
# Downloading and importing data
penguins_raw <- map_dfr(uris, read_csv) %>% 
  clean_names() %>%
  mutate(species_short = word(species, 1)) %>%
  select(species_short,
         island,
         culmen_length_mm,
         culmen_depth_mm,
         flipper_length_mm,
         body_mass_g,
         sex) %>%
  rename(species = species_short)
```


```{r}
library(ggplot2)
p <- ggplot(penguins_raw, aes(x = culmen_length_mm, y = culmen_depth_mm, color = species, fill = species)) +
  geom_point() +
  geom_smooth(method = "lm", formula = "y ~ x", alpha = 0.3) +
  scale_color_viridis_d() +
  scale_fill_viridis_d() +
  labs(x = "Culmen Length (mm)", y = "Culmen depth (mm)", fill = "Species", color = "Species")
p
```

2. Besides, it's better if you know how to create a R Markdown document and you know how to include R code in it (hint : in a _chunk_!). If not, start [here](https://bookdown.org/yihui/rmarkdown/).

# Making your own `ggplot2` theme

`ggplot2` theme manages how your graphic looks like. All elemements can be changed through the [`theme()`](https://ggplot2.tidyverse.org/reference/theme.html) function but there are also pre-configured one. The default theme used by `ggplot2` is `theme_gray()` but I often switch for `theme_bw()` (for _black and white_).


```{r animation.hook="gifski", interval = 2}
p + labs(title = "Default theme")
p + theme_bw() + labs(title = "With theme_bw()")
```   

If you always use the same modifications with `theme()` function, I highly suggest that you create your own theme. Here is a customised one :

```{r}
theme_ben <- function(base_size = 14) {
  theme_bw(base_size = base_size) %+replace%
    theme(
      # The whole figure
      plot.title = element_text(size = rel(1), face = "bold", margin = margin(0,0,5,0), hjust = 0),
      # Graphic area
      panel.grid.minor = element_blank(),
      panel.border = element_blank(),
      # Axis
      axis.title = element_text(size = rel(0.85), face = "bold"),
      axis.text = element_text(size = rel(0.70), face = "bold"),
      axis.line = element_line(color = "black", arrow = arrow(length = unit(0.3, "lines"), type = "closed")),
      # Legend
      legend.title = element_text(size = rel(0.85), face = "bold"),
      legend.text = element_text(size = rel(0.70), face = "bold"),
      legend.key = element_rect(fill = "transparent", colour = NA),
      legend.key.size = unit(1.5, "lines"),
      legend.background = element_rect(fill = "transparent", colour = NA),
      # Facet strips
      strip.background = element_rect(fill = "#17252D", color = "#17252D"),
      strip.text = element_text(size = rel(0.85), face = "bold", color = "white", margin = margin(5,0,5,0))
    )
}
```

And the results compared to the default one :

```{r animation.hook="gifski", interval = 2}
p + labs(title = "Default theme")
p + theme_ben() + labs(title = "With the customized one")
``` 

Building a customised theme is done by creating a R function where a pre-configured theme is used but some elements are modified with the `theme()` function. The tip is in the use of `%+replace%` in place of the classic `+`. Unlike the latter, `%+replace%` doesn't only update elements of a theme but replaces them entirely.

Now you just have to let your creativity flows. There are lots of editable elements so the customisation is pretty much limitless. You can find the list of elements in the webpage of the [`theme()`](https://ggplot2.tidyverse.org/reference/theme.html) function. In the example, the modified elements are for the whole figure (`plot.*` elements), the grapihic area (`pnael.*`), the axes (`axis.*`), the legend (`legend.*`) and the graphics labels (`strip.*`) when [facetting](https://ggplot2-book.org/facet.html) is used.

In my opinion, axis and legends are essential elements so my choices go towards highlighted them through their relative size using `rel()`, function which return a proportion the base size (`base_size`), and bolding theme (`face = "bold"`).

Customisation of a `ggplot2` theme can first be hard work but if you are going to often use the same configurations, it's worth it. There are plenty ressources on the web. You can also [contact me](https://benjaminlouis-stat.fr/contact/), I'll be glad to help.

You can now insert your theme in a _chunk_ at the beginning of your R Markdown document to use it all along. You can also create a R package with your theme, among others, and load this package.

# Your `ggplot2` figure in R Markdown

## Chunk options for figures

Inserting R code and its results in a R Markdown document is possible through utilisation of a _chunk_ which can take several options.

````markdown
`r ''````{r option1 = valeur_option1, option2 = valeur_option2}
# Le code R ici
`r ''````
````

Some of these options are specifics to figures made with R :

+ Options linked to the size of these figures when produced by R

+ Options linked to the size of these figures in the final document


## Size options of figures produced by R

Options `fig.width` and `fig.height` enable to set width and height of R produced figures. The default value is set to 7 (inches). When I play with these options, I prefer using only one of them (`fig.width`) in association with another one, `fig.asp`, which sets the _height-to-width_ ratio of the figure. It's easier in my mind to play with this ratio than to give a width and a height separatetly. The default value of `fig.asp` is `NULL` but I happen to often set it to $0.8$ because it corresponds to the expected result most of the time.

Size options of figures produced by R have consequences on relative sizes of elements in these figures. For a `ggplot2` figure, these elements will remain to the size defined in the used theme, whatever the choosen chunk option values. Therefore a huge size can lead to a very small text and vice versa.


<div class = "grid-container">
<div class = "row">
<div class = "col-6">

__`fig.width = 3` - figure elements too big__

````markdown
`r ''````{r fig.asp = 0.8, fig.width = 3}
p + theme_ben()
`r ''````
````

```{r echo = FALSE, fig.asp = 0.8, fig.width = 3, out.width = "95%"}
p + theme_ben()
```

</div>
<div class = "col-6">

__`fig.width = 10` - figure elements too small__


````markdown
`r ''````{r fig.asp = 0.8, fig.width = 10}
p + theme_ben()
`r ''````
````

```{r echo = FALSE, fig.asp = 0.8, fig.width = 10, out.width = "95%"}
p + theme_ben()
```

</div>
</div>
</div>

To find the result you like, you'll need to combine sizes set in your theme and set in the chunk options. With my customised theme, the default size (`7`) looks good to me.


````markdown
`r ''````{r fig.asp = 0.8, fig.width = 7}
p + theme_ben()
`r ''````
````

```{r echo = FALSE, fig.asp = 0.8, fig.width = 7}
p + theme_ben()
```


When texts axis are longer or when figures is overloaded, you can choose bigger size (`8` or `9`) to relatively reduce the figure elements. It's worth noting that for the text sizes, you can also modify the base size in your theme to obtain similar figures.


<div class = "grid-container">
<div class = "row">
<div class = "col-6">

__`fig.width = 9` - `base_size = 14` (default)__

````markdown
`r ''````{r fig.asp = 0.8, fig.width = 9}
p + theme_ben(base_size = 14)
`r ''````
````

```{r echo = FALSE, fig.asp = 0.8, fig.width = 9, out.width = "95%"}
p + theme_ben(base_size = 14)
```

</div>
<div class = "col-6">

__`fig.width = 7` (default) - `base_size = 12`__


````markdown
`r ''````{r fig.asp = 0.8, fig.width = 7}
p + theme_ben(base_size = 12)
`r ''````
````

```{r echo = FALSE, fig.asp = 0.8, fig.width = 7, out.width = "95%"}
p + theme_ben(base_size = 12)
```

</div>
</div>
</div>

## Size of figures in the final document

Figures made with R in a R Markdown document are exported (by default in png format) and then inserted in the final rendered document. Options `out.width` and `out.height` enable to choose the final size of the figure.

I nearly never need to rescale _height-to-width_ ratio after the figures were produced with R and this ratio is kept if you modify only one of both options therefore I only use `out.width`. For me, ercentages seem to be a good way to define the size of output figures. For example with a size set to 100% :


````markdown
`r ''````{r fig.asp = 0.8, fig.width = 7, out.width = "100%}
p + theme_ben()
`r ''````
````

```{r echo = FALSE, fig.asp = 0.8, fig.width = 7, out.width = "100%"}
p + theme_ben()
```

and a size set to 50% :

````markdown
`r ''````{r fig.asp = 0.8, fig.width = 7, out.width = "50%}
p + theme_ben()
`r ''````
````

```{r echo = FALSE, fig.asp = 0.8, fig.width = 7, out.width = "50%"}
p + theme_ben()
```


You can see here that the relative size of elements in the figure are unchanged. The figure is just more or less big. 

# Don't repeat yourself

It's quite annoying to write several time the exact same thing. Here, for every chunk with a `ggplot2` figure, you need add a line to specify your newly customised theme and you have to configure options in each chunk. Don't worry, solutions to deal with your (my!) lazyness exist.


## Changing default `ggplot2` theme

You can change the default `ggplot2` theme with the `theme_set()` function. You just have to write this line after creating your own theme (or loading the package with the theme you want) :
:

```{r eval = FALSE}
theme_set(theme_ben()) # change theme_ben() with the wanted theme
```

##  Changing default values of chunk options

You can also change default values of chunk options by writing this at the beginning of your R Markdown document :

````markdown
`r ''````{r setup, include=FALSE}
knitr::opts_chunk$set(
 fig.width = 6,
 fig.asp = 0.8,
 out.width = "80%"
)
`r ''````
````

These values will be applied for all chunks unless you specify other value in a chunk locally. You can set values often used (which differ from the default one) and avoid repeating them for each chunk.

## What can be the beginning of your R Markdown document

With all these tips, here I show an example of the beginning of a R Markdown document (HTML output format) with customised theme and chunk options :


````markdown
---
title: "My awesome title"
author: "Benjamin Louis"
date: "25/06/2020"
output: html_document
---


`r ''````{r setup, include=FALSE}
# Chunk options
knitr::opts_chunk$set(
 fig.width = 6,
 fig.asp = 0.8,
 out.width = "80%"
)
`r ''````

`r ''````{r theme_ggplot2, echo = FALSE}
# Creating a ggplot2 theme
library(ggplot)
theme_ben <- function(base_size = 14) {
  theme_bw(base_size = base_size) %+replace%
    theme(
      # changed theme options
    )
}
# Changing the default theme
theme_set(theme_ben())
`r ''````
````

From now on, you can modify what you need to create your `ggplot2` theme et give chunk options you like better. Of course, these options are not limited to figures produced by R, you can look at [this webpage](https://yihui.org/knitr/options/) to discover others.

I hope this post will help you write report you like and feel free to share your tips in the comments section !

<div class = "smaller">

#### Data citations

> **Palmer penguins data where originally published in** : [**Gorman KB, Williams TD, Fraser WR** (2014) Ecological Sexual Dimorphism and > Environmental Variability within a Community of Antarctic Penguins (Genus *Pygoscelis*). PLoS ONE 9(3): e90081. > doi:10.1371/journal.pone.0090081](https://journals.plos.org/plosone/article?id=10.1371/journal.pone.0090081)

</div>
